---
title: "index"
authors: "Kyle, Santiago"
format: html
editor: visual
---

```{r, message= FALSE}
library(tidyverse)
library(readxl)
data <- read_excel("/Users/KyleK/Desktop/DS/assignment04root/assignment04/Data/journalists_Sep22.xlsx")

view(data)
#dataset doesn't have a URL so the xlsx file must be downloaded.
```

# Assignment 04: Data Visualization, Quarto, and Git

## Visualization #1: Count of Events by Event Type

```{r}
library(ggplot2)
# Calculate counts for each disorder type.
data_counts <- data %>%
  group_by(DISORDER_TYPE) %>%
  summarise(count = n())

data_counts %>%
  ggplot(aes(y = DISORDER_TYPE, x = count)) +
  geom_bar(stat = "identity", width = 0.5, fill = "blue") +
  geom_text(aes(label = count), vjust = -0.1, hjust = -0.5, size = 3, color = "black") +
  theme(
    plot.title = element_text(face = "bold"),
    plot.subtitle = element_text(face = "italic")
  ) + 
  labs(title = "Number of Disorders by Type", subtitle = "1997 - 2023", y = "Count", x = "Disorder Type") +
  xlim(0, 15000) #modified x-axis limits to fit count labels onto plot. 

```

### Interpretation:

Each of these instances involves journalists and media workers: journalists, media editors, camera operators and photographers, news reporters/broadcasters, etc. The graph shows the number of disorders recorded in the dataset by disorder type. Demonstrations are the most common event, with over 10000 instances recorded, followed by political violence, with over 4500 instances recorded. Political violence; Demonstrations are the least common event, with under 200 instances recorded, followed by strategic developments with under 800 instances recorded. 

## Visualization #2: Fatalities by Disorder Type

```{r}
# Calculate the sum of FATALITIES for each EVENT_TYPE and DISORDER_TYPE
event_counts <- data |>
  group_by(DISORDER_TYPE, EVENT_TYPE) |>
  summarise(event_fat = sum(FATALITIES))

event_counts |>
  ggplot(aes(x = DISORDER_TYPE, y = event_fat, fill = EVENT_TYPE)) +
  geom_col() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  geom_text(aes(label = event_fat), position = position_stack(vjust = 0.5), size = 3, color = "black") + 
  #labels showing the sum of FATALITIES for each EVENT_TYPE
  labs(title = "Number of Fatalities by Disorder Type", subtitle = "1997 - 2023", x = "Disorder Type", y = "Fatalities", fill = "Event Type") +
  theme(plot.title = element_text(face = "bold"), plot.subtitle = element_text(face = "italic"))

```

### Interpretation:

The data set also includes events, which are sub categories of disorders. This graph displays the number of fatalities by disorder type and their respective subcategories. Political violence is the type of disorder with the highest number of recorded fatalities, with explosions/remote violence and violence against civilians being the deadliest sub categories.

## Visualization #3: Fatalities by Region

```{r}
data%>%
  ggplot() +
  geom_point(mapping = aes(x=FATALITIES, y = REGION), color = "red", alpha = 0.5) +
  labs(title = "Number of Fatalities by Region", subtitle = "1997 - 2023", x = "Fatalities", y = "Regions") +
  theme(plot.title = element_text(face = "bold"), plot.subtitle = element_text(face = "italic"))
```

### Interpretation:

This graph illustrates the number of fatalities for each individual observation across different regions. The majority of observations are low-fatality incidents. The middle east displays a high density of incidents with less than 25 fatalities, whereas Eastern and Northern africa are the only regions where an incident with 100 or more fatalities took place.

## Visualization #4: Fatalities by Region Over a Period of Time

```{r, message= FALSE, warning= FALSE}
data_sum <- data %>%
  group_by(YEAR, REGION) %>%
  summarize(Total_Fatalities = sum(FATALITIES))

regions_to_include <- c("Eastern Africa", "Northern Africa", "Middle East", "South Asia")
data_filtered <- data_sum %>%
  filter(REGION %in% regions_to_include)

data_filtered |>
  ggplot(aes(x = YEAR, y = Total_Fatalities, color = REGION)) +
  geom_line(size = 1.5, linetype = "solid") +
  geom_text(aes(label = Total_Fatalities), vjust = -0.5, hjust = 0.5, size = 3) +
  theme_minimal() +
  labs(x = "Year", y = "Fatalities per Year", title = "Fatalities Over Time by Region", subtitle = "1997 - 2023", color = "Region") +
  theme(plot.title = element_text(face = "bold"), plot.subtitle = element_text(face = "italic"))
```

### Interpretation:

The graph displays the fluctuations in fatalities per year in the Middle East, Northern Africa and Eastern Africa. All 3 regions had their deadliest years after 2015, although records for the Middle East prior to 2015 are absent from the data set. Northern Africa has the smallest peak out of all 3 regions, reaching 100 fatalities in 2019, whereas eastern Africa had relatively low fatalities up to 2020, and proceeded to peak at over 125 fatalities in 2022.
